---
title: "Sample Size Calcualtion for the Cochran-Mantel-Haenszel Test"
author:
  "Paul W. Egeler, M.S., GStat"
date: "`r Sys.Date()`"
output: word_document
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Sample Size Calcualtion for the Cochran-Mantel-Haenszel Test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<!--
May someday want output:
  pdf_document
  rmarkdown::html_vignette
  word_document
-->

## Background

The first method to estimate sample size for the CMH test was introduced by @Gail1973 using a least squares method. @Munoz1984 later produced a sample size estimator useful in the special case when both margins are fixed. However, this is often not the case in retrospective case-control designs. As such, @Woolson1986 introduced a calculation using the weighted difference between two binomial distributions. This was corroborated with @Wittes1987 in their paper a year later. Later still, @Nam1992 improved upon that calculation by introducing a coefficient to address continuity corrected CMH tests.

The methods included in the package are as follows:

`cc.binomial`
  : Continuity corrected weighted binomial sample size estimator described
  in the @Nam1992. This uses the weighted binomial sample size
  calculation described in @Woolson1986 but is enhanced
  for use with the continuity corrected Cochran's test.

`binomial`
  : This sample size calculation is designed for case-control studies
  where one margin is fixed. The method is "based on the
  Cochran-Mantel-Haenszel statistic expressed as a weighted difference in
  binomial proportions."
  Described in the @Woolson1986.

`hypergeometric`
  : Described in @Munoz1984. @Woolson1986
  state that this is the appropriate sample size estimator when both
  margins are fixed.

`least.squares`
  : Described in @Gail1973.

`unstratified`
  : The use of this method ignores stratification of the $J$ tables,
  essentially turning the data into a simple 2 $\times$ 2 table for
  calculation of sample size.

## The Research Question

Starting with the example first used by @Woolson1986 and then revisited by @Nam1992, we will explore the usage of the central function of the `samplesizeCMH`{.r} package, the `samplesize.cmh()`{.r} function.

This was a case--control study which looked at whether there was an association between colon cancer and chlorinated drinking water among males in Iowa. There were four age categories, 20--54 years, 55--69 years, 70--79 years, and 80--84 years; therefore $J$ = 4. The relative proportion of each age group to the total sample is as follows: $t_1$ = 0.10, $t_2$ = 0.40, $t_3$ = 0.35, and $t_4$ = 0.15. The National Collaborative Bladder Cancer Study showed that exposure rate from controls for each age stratum, denoted by $\pi_{2j}$ (where $j$ is the stratum), was $\pi_{21}$ = 0.75, $\pi_{22}$ = 0.70, $\pi_{23}$ = 0.65, and $\pi_{24}$ = 0.60. Level of significance was set at 0.05 and desired power was 90% with an effect size of 3. They matched equal number of cases and controls in each stratum, *i.e.*, $s_1=s_2=s_3=s_4=$ 0.5.

## Calculating Sample Size

### Methods

We will go over several calculation methods to compare the usage of the `samplesize.cmh()`{.r} function for different data sets and experimental designs.

#### Unstratified

Suppose that the researchers first decided to ignore the stratification variable in their sample size calculation. The researcher could use the `method = "unstratified"`{.r} argument. We see that the researcher has put in the control exposure proportions as a vector (`p2`), the effect size has been given (`theta`), and a vector of relative stratum sizes has been assigned (`t`). Since there are equal numbers of cases and controls for each group, the `s` parameter is not required. Only `p1` and `p2` *or* one of the proportions and a `theta` are required to carry out the calculation. The number of strata ($J$) is inferred from the maximum vector length of `p1`, `p2`, or `theta`.

```{r unstratified}
library(samplesizeCMH)

sample_size_unstratified <- samplesize.cmh(
  p2 = c(0.75,0.70,0.65,0.60),
  theta = 3,
  power = 0.9,
  t = c(0.10,0.40,0.35,0.15),
  alternative = "one",
  method = "unstratified"
)

sample_size_unstratified
```

#### Weighted Difference between Binomial Distributions

The researcher soon realized that failing to consider the confounding variable tends to lead to an underestimation of sample size. Adjusting the code above, an alteration is made to the `method` parameter, now using the calculation introduced by @Woolson1986. Note that the `"binomial"`{.r} option has been abbreviated.

```{r binomial}
# Uncorrected sample size estimate first introduced
# by Woolson and others in 1986
sample_size_binomial <- samplesize.cmh(
  p2 = c(0.75,0.70,0.65,0.60),
  theta = 3,
  power = 0.9,
  t = c(0.10,0.40,0.35,0.15),
  alternative = "one",
  method = "bin"
)

sample_size_binomial
```

In this case, the sample size has decreased slightly with this new calculation method. Meaning that the unstratified calcuation slightly overestimated sample size with respect to the formula proposed by @Woolson1986.

#### Continuity Corrected Estimate

The researcher now realizes that a continuity correction would be appropriate when perfroming the CMH test (the default of the `mantaelhaen.test()`{.r} is `correct = TRUE`{.r}). Another adjustment is made to the `method` parameter in the code below to take continuity correction into account.

```{r cc}
# Continuity corrected sample size estimate added by Nam
sample_size_corrected <- samplesize.cmh(
  p2 = c(0.75,0.70,0.65,0.60),
  theta = 3,
  power = 0.9,
  t = c(0.10,0.40,0.35,0.15),
  alternative = "one",
  method = "cc.bin"
)

sample_size_corrected
```

We see that the `N.exact` is indeed equal to that which is reported in the paper by @Nam1992.

```{r cc2}
sample_size_corrected$N.exact
```

#### Unequal Numbers of Cases and Controls

Consider now that the researcher performed some preparatory work and realized that the number of cases available is about $\frac{1}{3}$ the number of available controls. Wanting to leverage all of the available data, an adjustment was be made to the sample size calculation performed earlier by including the `s` parameter.

```{r s}
samplesize.cmh(
  p2 = c(0.75,0.70,0.65,0.60),
  s = 1/3,
  theta = 3,
  power = 0.9,
  t = c(0.10,0.40,0.35,0.15),
  alternative = "one",
  method = "cc.bin"
)
```


## A Closer Look at a Partial Table

Let's take a moment to look at the partial table of stratum $j$ = 1. Given the information above, we can reconstruct this table to better understand the question at hand. We can also use this as an opportunity to use some of the peripheral functions included in the `samplesizeCMH`{.r} package.

The proportion of controls exposed to chlorinated drinking water was $\pi_{21}$ = 0.75. It follows then that the complementary column proportion is 0.25. The table below shows column percents.


|           |Case | Control |
------------|:---:|:-------:|
Exposed     | ?   | 0.75
Not Exposed | ?   | 0.25

Using the `effect.size()`{.r} function, we can use the expected exposure rate of control to estimate expected exposure rate of cases.

```{r effect-size}
effect.size(0.75,3)
```

We can now use that to fill in the remainder of our table.

|           |Case | Control |
------------|:---:|:-------:|
Exposed     | 0.90| 0.75
Not Exposed | 0.10| 0.25

To determine the odds of exposure for the two groups, we can use `prop2odds()`{.r}.

```{r prop2odds}
# Control
prop2odds(0.75)

# Case
prop2odds(0.9)
```

We can also use either row or column proportions to calculate the odds ratio for a 2 $\times$ 2 table.

```{r props2theta}
props2theta(0.90,0.75)
```

There are several other peripheral functions in this package to interconvert between proportions, odds, relative risk, and odds ratios. Use `help("odds.and.proportions", "samplesizeCMH")`{.r} to find out more.

## Conclusion

## References
